---
title: "T2D_Proteomics_Houser"
author: "Madelyn Houser"
date: "6/7/2022"
output: 
  html_document:
    toc: yes
  word_document:
    toc: yes
---
  
# Load packages
  
```{r setup, message=FALSE}
library("tidyverse")
library("MetaboAnalystR")
library("qs")
library("limma")
library("pheatmap")
library("RColorBrewer")
library("fgsea")
library("data.table")
```
  
  
# Data Processing
  
## Load data 
```{r FeatureTable}
FT <- read.delim("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/proteomics_analysis/proteins.txt")

Key <- read.csv("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/SampleKey.csv")
```
  
  
## Data wrangling
  
### Format data frame
```{r IDs}
# Combine mz and time into a single variable separated by underscore
FT$mz_rt <- paste(FT$mz, FT$time, sep="_")
FT <- FT %>% dplyr::select(mz_rt, C1, C2, C3, D1, D2, D3)

# Keep only samples included in proteomics analysis in sample key
Key$CodedID <- c("C1", NA, NA, "C2", "C3", "D2", "D3", "D1", NA, NA)
Key <- Key %>% filter(!is.na(Key$CodedID)) %>% arrange(CodedID)

# Replace coded IDs with sample IDs
names(FT) <- c("mz_rt", Key$IIDPID)
```
  
### Format sample key
```{r key, message=FALSE}
samkey <- read_csv("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/proteomics_analysis/proteins_key.csv")
samkey$mz_rt <- paste(samkey$mz, samkey$time, sep="_")
```

  
  
### Check for missing data
```{r}
# Count missing data
sum(is.na(FT))
```

**Missing data count in expression table = `r sum(is.na(FT))`**   
  
  
### Format expression table for MetaboAnalyst
```{r formatFT}
# MetaboAnalyst expects a expression table in which:
## the first column contains mz and retention times for peaks separated by an underscore
## the other columns contain peak intensities for each sample
## column names reflect sample IDs
## the first row of the spreadsheet (below headers) contains labels indicating to which group each sample belongs


# Add group labels to expression table

## Cut and paste expression table down one row to leave space for a row of group labels
FT[2:(nrow(FT)+1),] <- FT[1:nrow(FT),]
FT[1,1] <- "Label"

## Replace first row with group labels
FT[1,2:ncol(FT)] <- Key$Group
```
  
  
## Save formatted expression tables  
```{r saveFTs}
# Save the formatted expression tables
write_csv(FT, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/ExpressionTable.csv")
```
  
  
## Pre-filtering
  
### Examine sample numbers in each group to determine appropriate filtering thresholds.
```{r groupnumbers}
# Check that we have enough samples in each group for our variable of interest to set our filtering thresholds appropriately in the next steps
sum(Key$Group=="Diabetic")
sum(Key$Group=="Control")
sum(Key$Group=="Diabetic") + sum(Key$Group=="Control")

# Determine which group has the smaller sample size
if((sum(Key$Group=="Diabetic")/(sum(Key$Group=="Diabetic") + sum(Key$Group=="Control"))) < (sum(Key$Group=="Control")/(sum(Key$Group=="Diabetic") + sum(Key$Group=="Control")))){
  small <- sum(Key$Group=="Diabetic")/(sum(Key$Group=="Diabetic") + sum(Key$Group=="Control"))
} else {
  small <- sum(Key$Group=="Control")/(sum(Key$Group=="Diabetic") + sum(Key$Group=="Control"))
}


# If it is possible to have a metabolite detected in 2/3 of one group and no samples of the other group and still not be detected in 30% of samples:
if(((2/3)*small)<0.3){
  ## then set a value that is lower than 2/3*(smallest group size/total group size) as the filtering threshold
  thresh <- floor(((2/3)*small)*100)/100
} else {
  ## otherwise set the filtering threshold to 0.3
  thresh <- 0.3
}
```
  
**The proportion of diabetic subjects is `r (sum(Key$Group=="Diabetic"))/(sum(Key$Group=="Diabetic") + sum(Key$Group=="Control"))` (`r sum(Key$Group=="Diabetic")` out of `r sum(Key$Group=="Diabetic") + sum(Key$Group=="Control")`).**  
**The proportion of control subjects is `r (sum(Key$Group=="Control"))/(sum(Key$Group=="Diabetic") + sum(Key$Group=="Control"))` (`r sum(Key$Group=="Control")` out of `r sum(Key$Group=="Diabetic") + sum(Key$Group=="Control")`).**  
**Is it possible for a protein to be found in 2/3 of the samples in one group but still in less than 30% of the total samples? `r ((2/3)*small) < 0.3`**  
**We will set our filtering thresholds accordingly.**
  
### Filter proteins based on prevalence in samples
  
#### Filter based on prevalence in all samples
```{r filterfeatures}
# Keep only features found in at least 30% of total samples

## Create variables reflecting the number of samples in each group in which the feature was not found
FTf <- FT %>% mutate(zeros = (rowSums(FT == 0)))

## Keep only samples that meet designated criterion
FTf <- FTf[FTf$zeros <= .7*nrow(Key), ]
```
  
#### Filter based on prevalence in each group
```{r filterfeatures2}
# Keep only features found in at least 2/3 of samples in an group

## Create variables reflecting the proportion of samples in each group in which the feature was not found
FTf <- FTf %>% mutate(diabeticzeros = (rowSums(FTf[,FTf[1,]=="Diabetic"] == 0))/length(FTf[,FTf[1,]=="Diabetic"]))
FTf <- FTf %>% mutate(controlzeros = (rowSums(FTf[,FTf[1,]=="Control"] == 0))/length(FTf[,FTf[1,]=="Control"]))

## Keep only samples that meet designated criterion
FTf2 <- FTf[FTf$diabeticzeros <= (1/3) | FTf$controlzeros <= (1/3), ]

## Remove created variables from data set
FTf2 <- FTf2 %>% dplyr::select(-diabeticzeros, -controlzeros, -zeros)
```
  
**Before filtering, `r nrow(FT)-1` proteins were identified.**  
**After filtering, `r nrow(FTf2)-1` proteins were retained.**
  
  
## Missingness
  
### Convert zeros to NA
```{r zeroNA}
FTf2[FTf2==0] <- NA
```
  
**A total of `r sum(is.na(FTf2))` (`r round(sum(is.na(FTf2))/(nrow(FTf2)*(ncol(FTf2)-1))*100, digits=2)`%) missing values were detected in the dataset.**
  
### Impute missing data
```{r imputemissing}
# Replace missing values with 1/2 minimum relative peak intensity detected in dataset
## Identify minimum
minFTf2 <- as.numeric(min(as.matrix(FTf2[2:nrow(FTf2),2:length(FTf2)]), na.rm=TRUE))

## Replace missing values with 1/2 minima
FTf2[is.na(FTf2)] <- minFTf2/2
```
  
  
## Save clean data
```{r savecleandata}
# Save the formatted expression tables
write_csv(FTf2, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/ExpressionTable_prefiltered.csv")
```
  
  
  
  
  
# MetaboAnalystR Analysis
  
## Initialization
```{r initialize}
# Create the mSet Object (struct), specifying that the data to be uploaded is a peak table ("pktable") and that statistical analysis will be performed ("stat").
mSet<-InitDataObjects("pktable", "stat", FALSE)
```
  
```{r readpktbl}
# Read in the filtered peak list
mSet<-Read.TextData(mSet, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/ExpressionTable_prefiltered.csv", "colu", "disc")
```
  
```{r sanitycheck}
mSet<-SanityCheckData(mSet)
```
  
```{r MRminrep}
# Perform data processing - Minimum Value Replacing
## By default, this replaces NAs with 1/5 of the minimum value of each feature
mSet <- ReplaceMin(mSet)
# We did this imputation manually above, and running it should have no effect on the data because we have no missing values, but for some reason, this step is needed to make the next steps function.

# Remove features containing a user-defined % cut-off of missing values
# mSet <- RemoveMissingPercent(mSet, percent=1)
# We did this manually, too
```
  
  
## Normalization and Transformation
```{r filternorm}
# Perform data processing - Variable Filtering and Normalization
# mSet<-FilterVariable(mSet, filter="none", qcFilter="F", rsd=25)
mSet<-PreparePrenormData(mSet)
# No normalization, log10 transformation, no scaling
mSet<-Normalization(mSet, "NULL", "LogNorm", "NULL")
```
  
```{r plotnorm}
mSet<-PlotNormSummary(mSet, "norm_0_", "png", 72, width=NA)
mSet<-PlotSampleNormSummary(mSet, "snorm_0_", "png", 72, width=NA)
```
  
  
## Univariate Analyses
  
### Fold Change Analysis
```{r FC, message=FALSE}
# Perform fold-change analysis
mSet <- FC.Anal(mSet, cmp.type=1, paired=FALSE, fc.thresh = 1)
mSet <- PlotFC(mSet, "fc_0_", "png", dpi=300, width=NA)

FoldChange <- read_csv("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/fold_change.csv")
FCup <- FoldChange %>% filter(`log2(FC)`>=2)
FCdown <- FoldChange %>% filter(`log2(FC)`<=-2)

names(FoldChange)[1] <- "mz_rt"
FC <- left_join(samkey, FoldChange) %>% filter(!is.na(`Fold Change`))
write_csv(FC, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/fold_change_annot.csv")
```
  
#### `r nrow(FCup)` proteins are more than 2 fold (log2(FC)) more abundant in diabetic islets while `r nrow(FCdown)` proteins are  more than 2 fold less abundant.
  
  
### T-tests
```{r ttests}
# Get output for t-test analysis
# Raw data
mSet <- Ttests.Anal(mSet, nonpar=F, threshp=0.05, paired=FALSE, equal.var=TRUE, pvalType="raw", FALSE)
mSet <- PlotTT(mSet, "tt_0_raw_", "png", dpi=300, width=NA)
# Change name of saved .csv file
ttest <- read_csv("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/t_test.csv")
colnames(ttest)[1] <- "mz_rt"
colnames(ttest)[4] <- " -log10(p)"
ttest2 <- left_join(samkey, ttest) %>% filter(!is.na(FDR))
write_csv(ttest2, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/t_test_raw.csv")

# FDR corrected
mSet <- Ttests.Anal(mSet, nonpar=F, threshp=0.1, paired=FALSE, equal.var=TRUE, pvalType="fdr", FALSE)
mSet <- PlotTT(mSet, "tt_1_fdr_", "png", dpi=300, width=NA)
# Change name of saved .csv file
ttest2 <- read_csv("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/t_test.csv")
colnames(ttest2)[1] <- "mz_rt"
colnames(ttest2)[4] <- " -log10(p)"
ttest2 <- left_join(samkey, ttest2) %>% filter(!is.na(FDR))
write_csv(ttest2, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/t_test_fdr.csv")
```
  
#### The abundance of `r nrow(ttest2)` proteins is significantly (padj<0.1) altered in diabetic islets.
  
  
  
## Multivariate Analyses
  
### PCA
```{r PCA}
# Perform PCA
mSet <- PCA.Anal(mSet)
mSet <- PlotPCAPairSummary(mSet, "pca_pair_0_", "png", 300, width=NA, 5)
mSet <- PlotPCAScree(mSet, "pca_scree_0_", "png", 300, width=NA, 5)
mSet <- PlotPCA2DScore(mSet, "pca_score2d_0_", "png", 300, width=NA, 1,2,0.95,0,0)
mSet <- PlotPCALoading(mSet, "pca_loading_0_", "png", 300, width=NA, 1,2);
mSet <- PlotPCABiplot(mSet, "pca_biplot_0_", "png", 300, width=NA, 1,2)
mSet <- PlotPCA3DScoreImg(mSet, "pca_score3d_0_", "png", 300, width=NA, 1,2,3, 40)
```
  
  
  
# Limma analysis
```{r getnormdf}
# Import log-transformed dataset and format appropriately
dfnorm <- qread("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/complete_norm.qs")
dfnorm <- as.matrix(t(dfnorm))
```


```{r limma}
#Define the design vector
# data input should be matrix-like object with rows corresponding to genes and columns to samples
group = as.factor(Key$Group)
design = model.matrix(~0+group)
colnames(design) = gsub("group","",colnames(design))
#Make contrasts
x <- "Diabetic-Control"
contrast =  makeContrasts(contrasts=x,levels=design)
fit1 <- lmFit(dfnorm, design)
fit2 <- contrasts.fit(fit1,contrasts = contrast)
fit3 <- eBayes(fit2)
```
  
```{r limmares05}
# Get significant up/significant down/not significant output
sig <- decideTests(fit3, p.value=0.05, adjust.method = "none")
sig <- data.frame(sig)
sig$mz_rt <- rownames(sig)
names(sig)[1] <- "DiffAbund"

# Get results table
results <- topTable(fit3, number=200, p.value=0.05, adjust.method="none")

results$mz_rt <- rownames(results)
results2 <- left_join(samkey, results) %>% filter(!is.na(logFC))

# Update gene annotations with current symbols
results2$`Gene names` <- results2$`Gene names` %>% str_replace("BOLA2;BOLA2B;LOC101060252", "BOLA2B") %>% str_replace("C14orf159", "DGLUCY") %>% str_replace("C2orf47", "MAIP1") %>% str_replace("EPDR1;UCC1", "EPDR1") %>% str_replace("GUCY1B3", "GUCY1B1") %>% str_replace("HIST1H1B", "H1-5") %>% str_replace("MKI67IP", "NIFK") %>% str_replace("MKL2", "MRTFB") %>% str_replace("NIP30;FAM192A", "PSME3IP1") %>% str_replace("NRD1", "NRDC") %>% str_replace("RAB7L1", "RAB29") %>% str_replace("STX16;STX16-NPEPL1", "STX16") %>% str_replace("TCEB1", "ELOC") %>% str_replace("ZMIZ1;ZMIZ2", "ZMIZ1")

results2 <- results2 %>% mutate(log10p = -log10(P.Value)) 
write_csv(results2, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/limma_results05.csv")


FCup <- results2 %>% filter(logFC>0)
FCdown <- results2 %>% filter(logFC<0)

```
  
#### `r nrow(results2)` proteins differ significantly between diabetic and control islets. `r nrow(FCup)` proteins are significantly (p<0.05) more abundant in diabetic islets while `r nrow(FCdown)` proteins are significantly less abundant.
  
```{r limmaresFDR}
# Get significant up/significant down/not significant output
sigfdr <- decideTests(fit3, p.value=0.1, adjust.method = "fdr")
sigfdr <- data.frame(sigfdr)
sigfdr$mz_rt <- rownames(sigfdr)
names(sigfdr)[1] <- "DiffAbund"

# Get results table
resultsfdr <- topTable(fit3, number=100, p.value=0.1, adjust.method="fdr")

resultsfdr$mz_rt <- rownames(resultsfdr)
resultsfdr2 <- left_join(samkey, resultsfdr) %>% filter(!is.na(logFC))

resultsfdr2 <- resultsfdr2 %>% mutate(log10p = -log10(adj.P.Val)) 
write_csv(resultsfdr2, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/limma_resultsfdr_FDR1.csv")


FCupfdr <- resultsfdr2 %>% filter(logFC>0)
FCdownfdr <- resultsfdr2 %>% filter(logFC<0)

```
  
#### `r nrow(resultsfdr2)` proteins differ significantly between diabetic and control islets. `r nrow(FCupfdr)` proteins are significantly (padj<0.1) more abundant in diabetic islets while `r nrow(FCdownfdr)` proteins are significantly less abundant.
  
  
```{r methodcomp}
resultsfdr2$mz_rt %in% ttest2$mz_rt
```
  
#### All but one of the proteins identified by `limma` and by t-tests with FDR correction are the same.
  
  
## Volcano plot
```{r volcdat}
# Get limma p values
ps <- data.frame(fit3$p.value)
names(ps)[1] <- "p.value"
ps <- ps %>% mutate(padj.value=p.adjust(ps$p.value, method = "fdr", n = length(ps$p.value)))
ps$mz_rt <- rownames(ps)


# Merge fold change, raw p value, and differential abundance indicator data
volcdat <- left_join(FC, ps)
volcdat <- left_join(volcdat, sig)
volcdat$DiffAbund <- factor(volcdat$DiffAbund)
volcdat <- volcdat %>% mutate(log10p = -log10(p.value))

write_csv(volcdat, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/volcdat.csv")


# Merge fold change, FDR-adjusted p value, and differential abundance indicator data
volcdatFDR <- left_join(FC, ps)
volcdatFDR <- left_join(volcdatFDR, sigfdr)
volcdatFDR$DiffAbund <- factor(volcdatFDR$DiffAbund)
volcdatFDR <- volcdatFDR %>% mutate(log10p = -log10(padj.value))

write_csv(volcdatFDR, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/volcdat_fdr.csv")
```

```{r volcplot}
ggplot(volcdat, aes(x = `log2(FC)`, y = log10p, colour = DiffAbund, size=DiffAbund)) +
  geom_point(shape=19, alpha = 0.4) +
  scale_colour_manual(values=c("magenta", "black", "magenta")) +
  scale_size_manual(values=c(2, 1, 2)) +
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = -1) +
  geom_vline(xintercept = 1) +
  theme_classic() +
  theme(legend.position="none", panel.border=element_rect(fill=NA)) +
  labs(x ="log2 (FC)", y ="-log10(p)")

ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/volcano_limmaRaw.png", dpi=300)
```
  
```{r volcplotFDR}
ggplot(volcdatFDR, aes(x = `log2(FC)`, y = log10p, colour = DiffAbund, size=DiffAbund)) +
  geom_point(shape=19, alpha = 0.4) +
  scale_colour_manual(values=c("magenta", "black", "magenta")) +
  scale_size_manual(values=c(2, 1, 2)) +
  geom_hline(yintercept = 1) +
  geom_vline(xintercept = -1) +
  geom_vline(xintercept = 1) +
  theme_classic() +
  theme(legend.position="none", panel.border=element_rect(fill=NA)) +
  labs(x ="log2 (FC)", y ="-log10(padj)")

ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/volcano_limmaFDR.png", dpi=300)
```
  
  
## Heatmap
  
### Limma (raw)
```{r heatmapdata}
# Set gene names as row names for data set
dfhm <- data.frame(dfnorm)
dfhm$mz_rt <- rownames(dfhm)
dfhm <- right_join(samkey, dfhm) %>% filter(mz_rt %in% results2$mz_rt) %>% data.frame()
rownames(dfhm) <- dfhm$Gene.names
dfhm <- dfhm %>% dplyr::select(XIN460:AAJ2482) %>% as.matrix()

## Specify group annotations
grpann <- Key %>% dplyr::select(IIDPID, Group)
rownames(grpann) <- grpann$IIDPID
grpann <- grpann %>% dplyr::select(-IIDPID)
names(grpann) <- "Group"
```
  
```{r heatmap}
# Set colors for categories
annCol <- list(Group=c(Control="brown2", Diabetic="springgreen3"))
## TO specify above, add "annotation_colors=annCol," below

# Create heatmap with features identified by limma (p<0.05, not FDR-corrected)
hmttest <- pheatmap(dfhm, color=colorRampPalette(rev(brewer.pal(n=11, name="RdBu")))(100), scale="row", clustering_method="ward.D2", annotation_col=grpann, annotation_colors=annCol,  filename="C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/heatmap_limma.png", width=7, height=22)
```
   
### Limma (FDR-corrected)
```{r heatmapdatafdr}
# Set gene names as row names for data set
dfhm <- data.frame(dfnorm)
dfhm$mz_rt <- rownames(dfhm)
dfhm <- right_join(samkey, dfhm) %>% filter(mz_rt %in% resultsfdr2$mz_rt) %>% data.frame()
rownames(dfhm) <- dfhm$Gene.names
dfhm <- dfhm %>% dplyr::select(XIN460:AAJ2482) %>% as.matrix()

## Specify group annotations
grpann <- Key %>% select(IIDPID, Group)
rownames(grpann) <- grpann$IIDPID
grpann <- grpann %>% select(-IIDPID)
names(grpann) <- "Group"
```
  
```{r heatmapfdr}
# Create heatmap with features identified by limma (padj<0.1, FDR-corrected)
hmttest <- pheatmap(dfhm, color=colorRampPalette(rev(brewer.pal(n=11, name="RdBu")))(100), scale="row", clustering_method="ward.D2", annotation_col=grpann, annotation_colors=annCol,  filename="C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/heatmap_limmaFDR.png", width=6, height=9)
```

  
  
  
  
# Differences in individual proteins
  
## Get normalized H/L ratios for each sample for differentially abundant proteins  
```{r wranglesigprot}
# Get normalized H/L ratios for each sample for differentially abundant proteins
FTann <- right_join(samkey, FTf2) %>% filter(mz_rt %in% resultsfdr2$mz_rt)
sigprots <- FTann$`Protein names`
FTann <- FTann %>% dplyr::select(`Gene names`, XIN460:AAJ2482)

# Reformat data frame
FTannt <- data.frame(t(FTann))
names(FTannt) <- FTannt[1,]
FTannt <- FTannt[-1,]

# Add experimental group information to data frame
FTannt$Group <- Key$Group
```
  
```{r plotsigprot}
# Plot normalized H/L ratios for each differentially abundant protein
for(i in 1:(ncol(FTannt)-1)){
  p <- ggplot(aes(x=Group, y=FTannt[,i]), data=FTannt) +
    geom_jitter(width=0.08, size=3) +
    ylab(names(FTannt[i])) +
    ggtitle(sigprots[i]) +
    theme_bw() +
    theme(plot.title = element_text(size=12))
  print(p)
}
```
  
  
  
# Pathway Enrichment Analysis (GSEA)
  
```{r pathdf}
# Rank genes in entire data set according to product of log fold change and -log10(p)
# Remove data for which no gene symbol was identified
dfGSEA <- volcdat %>% mutate(rank = `log2(FC)`*log10p) %>% filter(`Gene names` != "---")

dfGSEA$`Gene names` <- dfGSEA$`Gene names` %>% str_replace_all("BOLA2;BOLA2B;LOC101060252", "BOLA2B") %>% str_replace_all("C14orf159", "DGLUCY") %>% str_replace_all("C2orf47", "MAIP1") %>% str_replace_all("EPDR1;UCC1", "EPDR1") %>% str_replace_all("GUCY1B3", "GUCY1B1") %>% str_replace_all("HIST1H1B", "H1-5") %>% str_replace_all("MKI67IP", "NIFK") %>% str_replace_all("MKL2", "MRTFB") %>% str_replace_all("NIP30;FAM192A", "PSME3IP1") %>% str_replace_all("NRD1", "NRDC") %>% str_replace_all("RAB7L1", "RAB29") %>% str_replace_all("STX16;STX16-NPEPL1", "STX16") %>% str_replace_all("TCEB1", "ELOC") %>% str_replace_all("ZMIZ1;ZMIZ2", "ZMIZ1")

```
  
`r nrow(dfGSEA)-(length(unique(dfGSEA$'Gene names')))` rows of the data set contain duplicated gene symbols.
  
```{r rmdupgene}
# Remove duplicated rows from the data set, keeping the row with the lowest p value
for(i in 1:length(unique(dfGSEA$`Gene names`))){
  if(dfGSEA$`Gene names`[i+1] == dfGSEA$`Gene names`[i]){
    ifelse(
      dfGSEA$p.value[i] <= dfGSEA$p.value[i+1],
      dfGSEA <- dfGSEA %>% filter(dfGSEA$mz_rt != dfGSEA$mz_rt[i+1]),
      dfGSEA <- dfGSEA %>% filter(dfGSEA$mz_rt != dfGSEA$mz_rt[i])
    )
  }
}

# Run again to remove duplicates remaining (occurs because originally genes were triplicated or quadruplicated)
for(i in 1:(length(unique(dfGSEA$`Gene names`))-1)){
  if(dfGSEA$`Gene names`[i+1] == dfGSEA$`Gene names`[i]){
    ifelse(
      dfGSEA$p.value[i] <= dfGSEA$p.value[i+1],
      dfGSEA <- dfGSEA %>% filter(dfGSEA$mz_rt != dfGSEA$mz_rt[i+1]),
      dfGSEA <- dfGSEA %>% filter(dfGSEA$mz_rt != dfGSEA$mz_rt[i])
    )
  }
}

nrow(dfGSEA) == length(unique(dfGSEA$`Gene names`))
```
  
  
```{r MoreSymbolCleanup}
# Cleanup gene symbol annotations for genes not identified as differing signicantly (p<0.05) by limma

dfGSEA$'Gene names' <- dfGSEA$`Gene names` %>% str_replace("NO66", "RIOX1") %>% str_replace("lRMP", "IRAG2") %>% str_replace("TMEM55A", "PIP4P2") %>% str_replace("QTRTD1", "QTRT2") %>% str_replace("FAM134C;LOC162427", "RETREG3") %>% str_replace("Em:AP000351.3;GSTT2;GSTT2B", "GSTT2") %>% str_replace("SSFA2", "ITPRID2") %>% str_replace("ATP5EP2;ATP5E", "ATP5F1E") %>% str_replace("FAM213B", "PRXL2B") %>% str_replace("WDR92;DKFZp434B156", "DNAAF10") %>% str_replace("FAM105B", "OTULIN") %>% str_replace("RP13-360B22.2;TMEM164", "TMEM164") %>% str_replace("HIST3H2BB", "H2BC26") %>% str_replace("TARSL2", "TARS3") %>% str_replace("KIAA1462", "JCAD") %>% str_replace("IFT140;DKFZp564L232", "IFT140") %>% str_replace("ADSSL1", "ADSS1") %>% str_replace("H1F0", "H1-0") %>% str_replace("SDPR", "CAVIN2") %>% str_replace("POLR2M;GCOM1;GCOM2", "GCOM1") %>% str_replace("KRT6C;KRT6A", "KRT6A") %>% str_replace("SMEK1", "PPP4R3A") %>% str_replace("PET112", "GATB") %>% str_replace("C10orf32", "BORCS7") %>% str_replace("FUK", "FCSK") %>% str_replace("PDPK1;PDPK2", "PDPK1") %>% str_replace("ATP5S", "DMAC2L") %>% str_replace("TENC1", "TNS2") %>% str_replace("WHSC1L1", "NSD3") %>% str_replace("ADCK3", "COQ8A") %>% str_replace("SMCR7L", "MIEF1") %>% str_replace("FAM129A", "NIBAN1")
```

  
  
```{r generank}
# Extract gene rank and list in decreasing order for GSEA
ranks <- as.numeric(dfGSEA$rank)
names(ranks) <- dfGSEA$`Gene names`
ranks <- sort(ranks, decreasing=TRUE)
ranks[1:10]
```
  
## Gene Ontology enrichment analysis
```{r gmt}
# Import .gmt file with pathway information by gene symbol (Gene sets derived from the GO Biological Process ontology)
## Obtained from GSEA 3-6-2023 http://www.gsea-msigdb.org/gsea/msigdb/collections.jsp
go <- gmtPathways("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/c5.go.bp.v2023.1.Hs.symbols.gmt")
```
  
```{r pathfgsea}
set.seed(06212022)

fgseaRes <- fgsea(pathways = go, 
                  stats = ranks,
                  minSize=6,
                  maxSize=500,
                  nproc=1)
```

  
```{r pathsel}
# write table
fwrite(fgseaRes, file="C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/fgseaRes_GO.txt", sep="\t", sep2=c("", " ", ""))

# read table
pathtab <- read.delim("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/fgseaRes_GO.txt")


# Select top 10 up and down-regulated pathways
topup <- pathtab %>% filter(ES > 0)
topup <- topup[order(topup$pval),]
topup <- topup[1:10,]

topdown <- pathtab %>% filter(ES < 0)
topdown <- topdown[order(topdown$pval),]
topdown <- topdown[1:10,]

top <- rbind(topup, rev(topdown))

# Clean up pathway names
top$pathway <- top$pathway %>% str_replace("GO.+?_", "") %>% str_replace_all("_", " ")

top <- top[order(top$pval),]
top <- top %>% filter(pval<=0.05) %>% filter(!is.na(pathway))


# Write cleaned table
write_csv(top, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/Top_GSEA_Pathways_GO.csv")
```
  
  
### Plot pathway analysis results
  
#### Significant pathways (padj < 0.1)
```{r sigpathgraph, fig.width=11, fig.height=3}
# Summarize pathway information and create negative log p value variable for graphing
pathg <- top %>% filter(padj<=0.1) 
pathg <- pathg %>% mutate(neglogpvalue=-log10(pval))

if(nrow(pathg)>=1){
  # Graph pathways by p value
  pathfig <- ggplot(pathg, aes(x=reorder(pathway, neglogpvalue), y=neglogpvalue)) +
    geom_bar(stat = "identity", fill="black") +
    coord_flip() +
    scale_x_discrete(name="Pathways Associated with Diabetes") +
    ylab("-log(p value)") +
    theme(axis.text.x = element_text(face="bold", size=10, angle=0),
          axis.text.y = element_text(face="bold", size=10, angle=0))

  pathfig

  ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/pathsigfig_GO.png")
}


```
  
#### Top 10 pathways
```{r pathgraph, fig.width=11, fig.height=6}
# Summarize pathway information and create negative log p value variable for graphing
pathg <- top[1:10,] 

pathg <- pathg %>% mutate(neglogpvalue=-log10(pval))

# Graph pathways by p value
pathfig <- ggplot(pathg, aes(x=reorder(pathway, neglogpvalue), y=neglogpvalue)) +
  geom_bar(stat = "identity", fill="black") +
  coord_flip() +
  scale_x_discrete(name="Pathways Associated with Diabetes") +
  ylab("-log(p value)") +
  theme(axis.text.x = element_text(face="bold", size=10, angle=0),
        axis.text.y = element_text(face="bold", size=10, angle=0))

pathfig

ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/TenPathfig_GO.png")
```
  
  
### Determine direction of change from one group to another
  
```{r pathdirection, message=FALSE}
# Extract mzs for overlapping features
genenames <- data.frame(str_split(as.character(top$leadingEdge[1])," "))
names(genenames) <- "LeadingEdge"

pathLE <- dfGSEA %>% filter(dfGSEA$'Gene names' %in% genenames$LeadingEdge) %>% rename(log2FC='log2(FC)')

# Create column just distinguishing metabolites
pathLE$feature <- c(1:nrow(pathLE))
```
  
```{r plotdirfeatures, fig.height=16, fig.width=10}
# Plot fold change for each metabolite in the identified pathways
fcplot <- ggplot(pathLE, aes(x=feature, y=log2FC)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  geom_text(size=3, label=pathLE$'Gene names', hjust=1.5) +
  scale_x_discrete(name="Organophosphate Metabolic Process Genes", labels=pathLE$'Gene names') +
  ylab("log2(Fold Change) in Diabetic Islets") +
  theme(axis.text.x = element_text(face="bold", size=10, angle=0),
        axis.text.y = element_text(face="bold", size=10, angle=0))
  
fcplot

ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/FCfeatures_organophosphate_met_process.png")
```
  
  
## KEGG enrichment analysis
```{r gmtKEGG}
# Import .gmt file with pathway information by gene symbol (Canonical Pathways gene sets derived from the KEGG pathway database)
## Obtained from GSEA 3-6-2023 http://www.gsea-msigdb.org/gsea/msigdb/collections.jsp
kegg <- gmtPathways("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/c2.cp.kegg.v2023.1.Hs.symbols.gmt")
```
  
```{r pathfgseaKEGG}
set.seed(06212022)

fgseaRes <- fgsea(pathways=kegg, 
                  stats = ranks,
                  minSize=6,
                  maxSize=500,
                  nproc=1)
```
  
```{r pathselKEGG}
# write table
fwrite(fgseaRes, file="C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/fgseaRes_KEGG.txt", sep="\t", sep2=c("", " ", ""))

# read table
pathtab <- read.delim("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/fgseaRes_KEGG.txt")


# Select top 10 up and down-regulated pathways
topup <- pathtab %>% filter(ES > 0)
topup <- topup[order(topup$pval),]
topup <- topup[1:10,]

topdown <- pathtab %>% filter(ES < 0)
topdown <- topdown[order(topdown$pval),]
topdown <- topdown[1:10,]

top <- rbind(topup, rev(topdown))

# Clean up pathway names
top$pathway <- top$pathway %>% str_replace("GO.+?_", "") %>% str_replace_all("_", " ")

top <- top[order(top$pval),]
top <- top %>% filter(pval<=0.05)

# Write cleaned table
write_csv(top, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/Top_GSEA_Pathways_KEGG.csv")
```
  
  
### Plot pathway analysis results
  
#### Significant pathways (padj < 0.1)
```{r sigpathgraphKEGG, fig.width=11, fig.height=3}
# Summarize pathway information and create negative log p value variable for graphing
pathg <- top %>% filter(padj<=0.1) 
pathg <- pathg %>% mutate(neglogpvalue=-log10(pval))

if(nrow(pathg)>=1){
  # Graph pathways by p value
  pathfig <- ggplot(pathg, aes(x=reorder(pathway, neglogpvalue), y=neglogpvalue)) +
    geom_bar(stat = "identity", fill="black") +
    coord_flip() +
    scale_x_discrete(name="Pathways Associated with Diabetes") +
    ylab("-log(p value)") +
    theme(axis.text.x = element_text(face="bold", size=10, angle=0),
          axis.text.y = element_text(face="bold", size=10, angle=0))

  pathfig

  ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/pathsigfig_KEGG.png")
}
```
  
#### Top 10 pathways
```{r pathgraphKEGG, fig.width=11, fig.height=5}
# Summarize pathway information and create negative log p value variable for graphing
pathg <- top[1:10,] 
pathg <- pathg %>% filter(!is.na(pathway))

pathg <- pathg %>% mutate(neglogpvalue=-log10(pval))

# Graph pathways by p value
pathfig <- ggplot(pathg, aes(x=reorder(pathway, neglogpvalue), y=neglogpvalue)) +
  geom_bar(stat = "identity", fill="black") +
  coord_flip() +
  scale_x_discrete(name="Pathways Associated with Diabetes") +
  ylab("-log(p value)") +
  theme(axis.text.x = element_text(face="bold", size=10, angle=0),
        axis.text.y = element_text(face="bold", size=10, angle=0))

pathfig

ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/TenPathfig_KEGG.png")
```
  
  
## Hallmark enrichment analysis
```{r gmtHM}
# Import .gmt file with pathway information by gene symbol (Hallmark gene sets)
## Obtained from GSEA 3-6-2023 http://www.gsea-msigdb.org/gsea/msigdb/collections.jsp
hm <- gmtPathways("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/h.all.v2023.1.Hs.symbols.gmt")
```
  
```{r pathfgseaHM}
set.seed(06212022)

fgseaRes <- fgsea(pathways=hm, 
                  stats = ranks,
                  minSize=6,
                  maxSize=500,
                  nproc=1)
```
  
```{r pathselHM}
# write table
fwrite(fgseaRes, file="C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/fgseaRes_HM.txt", sep="\t", sep2=c("", " ", ""))

# read table
pathtab <- read.delim("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/fgseaRes_HM.txt")


# Select top 10 up and down-regulated pathways
topup <- pathtab %>% filter(ES > 0)
topup <- topup[order(topup$pval),]
topup <- topup[1:10,]

topdown <- pathtab %>% filter(ES < 0)
topdown <- topdown[order(topdown$pval),]
topdown <- topdown[1:10,]

top <- rbind(topup, rev(topdown))

# Clean up pathway names
top$pathway <- top$pathway %>% str_replace("GO.+?_", "") %>% str_replace_all("_", " ")

top <- top[order(top$pval),]
top <- top %>% filter(pval<=0.05)

# Write cleaned table
write_csv(top, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/Top_GSEA_Pathways_HM.csv")
```
  
  
### Plot pathway analysis results
  
#### Significant pathways (padj < 0.1)
```{r sigpathgraphHM, fig.width=11, fig.height=3}
# Summarize pathway information and create negative log p value variable for graphing
pathg <- top %>% filter(padj<=0.1) 
pathg <- pathg %>% mutate(neglogpvalue=-log10(pval))

if(nrow(pathg)>=1){
  # Graph pathways by p value
  pathfig <- ggplot(pathg, aes(x=reorder(pathway, neglogpvalue), y=neglogpvalue)) +
    geom_bar(stat = "identity", fill="black") +
    coord_flip() +
    scale_x_discrete(name="Pathways Associated with Diabetes") +
    ylab("-log(p value)") +
    theme(axis.text.x = element_text(face="bold", size=10, angle=0),
          axis.text.y = element_text(face="bold", size=10, angle=0))

  pathfig

  ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/pathsigfig_HM.png")
}
```
  
#### Top 10 pathways
```{r pathgraphHM, fig.width=11, fig.height=3}
# Summarize pathway information and create negative log p value variable for graphing
pathg <- top[1:10,]
pathg <- pathg %>% filter(!is.na(pathway))

pathg <- pathg %>% mutate(neglogpvalue=-log10(pval))

# Graph pathways by p value
pathfig <- ggplot(pathg, aes(x=reorder(pathway, neglogpvalue), y=neglogpvalue)) +
  geom_bar(stat = "identity", fill="black") +
  coord_flip() +
  scale_x_discrete(name="Pathways Associated with Diabetes") +
  ylab("-log(p value)") +
  theme(axis.text.x = element_text(face="bold", size=10, angle=0),
        axis.text.y = element_text(face="bold", size=10, angle=0))

pathfig

ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/TenPathfig_HM.png")
```
  
  
## Cell type signature enrichment analysis
```{r gmtcell}
# Import .gmt file with pathway information by gene symbol (Gene sets that contain curated cluster markers for cell types identified in single-cell sequencing studies of human tissue)
## Obtained from GSEA 3-6-2023 http://www.gsea-msigdb.org/gsea/msigdb/collections.jsp
cts <- gmtPathways("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/c8.all.v2023.1.Hs.symbols.gmt")
```
  
```{r pathfgseacell}
set.seed(06212022)

fgseaRes <- fgsea(pathways=cts, 
                  stats = ranks,
                  minSize=6,
                  maxSize=500,
                  nproc=1)
```
  
```{r pathselcell}
# write table
fwrite(fgseaRes, file="C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/fgseaRes_cts.txt", sep="\t", sep2=c("", " ", ""))

# read table
pathtab <- read.delim("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/fgseaRes_cts.txt")


# Select top 10 up and down-regulated pathways
topup <- pathtab %>% filter(ES > 0)
topup <- topup[order(topup$pval),]
topup <- topup[1:10,]

topdown <- pathtab %>% filter(ES < 0)
topdown <- topdown[order(topdown$pval),]
topdown <- topdown[1:10,]

top <- rbind(topup, rev(topdown))

# Clean up pathway names
top$pathway <- top$pathway %>% str_replace("GO.+?_", "") %>% str_replace_all("_", " ")

top <- top[order(top$pval),]
top <- top %>% filter(pval<=0.05)

# Write cleaned table
write_csv(top, "C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/Top_GSEA_Pathways_cts.csv")
```
  
  
### Plot pathway analysis results
  
#### Significant pathways (padj < 0.1)
```{r sigpathgraphcell, fig.width=11, fig.height=3}
# Summarize pathway information and create negative log p value variable for graphing
pathg <- top %>% filter(padj<=0.1) 
pathg <- pathg %>% mutate(neglogpvalue=-log10(pval))

if(nrow(pathg)>=1){
  # Graph pathways by p value
  pathfig <- ggplot(pathg, aes(x=reorder(pathway, neglogpvalue), y=neglogpvalue)) +
    geom_bar(stat = "identity", fill="black") +
    coord_flip() +
    scale_x_discrete(name="Pathways Associated with Diabetes") +
    ylab("-log(p value)") +
    theme(axis.text.x = element_text(face="bold", size=10, angle=0),
          axis.text.y = element_text(face="bold", size=10, angle=0))

  pathfig

  ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/pathsigfig_cts.png")
}
```
  
#### Top 10 pathways
```{r pathgraphcell, fig.width=11, fig.height=6}
# Summarize pathway information and create negative log p value variable for graphing
pathg <- top[1:10,]
pathg <- pathg %>% filter(!is.na(pathway))

pathg <- pathg %>% mutate(neglogpvalue=-log10(pval))

# Graph pathways by p value
pathfig <- ggplot(pathg, aes(x=reorder(pathway, neglogpvalue), y=neglogpvalue)) +
  geom_bar(stat = "identity", fill="black") +
  coord_flip() +
  scale_x_discrete(name="Pathways Associated with Diabetes") +
  ylab("-log(p value)") +
  theme(axis.text.x = element_text(face="bold", size=10, angle=0),
        axis.text.y = element_text(face="bold", size=10, angle=0))

pathfig

ggsave("C:/Users/mecho/Documents/Hertzberg_Lab/T2D/Proteomics_for_Vicki/Analysis_Houser/Pathways/TenPathfig_cts.png")
```
  
  
  
  
```{r session}
sessionInfo()
```  
